Sample-Efficient Reinforcement Learning \\ with Implicitly Quantized Representations

### 2024-02-02 paper/main.tex
updated figs
added codebook stuff

\begin{algorithm}[tb]
   \caption{\our}
   \label{alg:main_alg}
   \renewcommand{\algorithmiccomment}[1]{\hfill\textcolor{gray}{\(\triangleright\) #1}}
\begin{algorithmic}
   \STATE {\bfseries Input:} Encoder $e_{\theta}$, transition $d_{\phi}$, critics $\{Q_{\psi_{1}}, Q_{\psi_{2}} \}$, policy $\pi_{\eta}$, learning rate $\alpha$, target network update rate $\tau$
   \FOR{$i$ {\bfseries to} $N_{\text{episodes}}$}
    \STATE Collect data in environment
    \STATE $\qquad \mathcal{D} \leftarrow \mathcal{D} \cup \{o_{t}, a_{t}, o_{t+1}, r_{t+1}\}^{T}_{t=0}$
    \FOR{$i=1$ {\bfseries to} $T$}
        \STATE Update representation
        \STATE $\qquad [\theta, \phi] \leftarrow [\theta, \phi] + \alpha \nabla \left( \mathcal{L}_{\text{rep}}(\theta, \phi; \mathcal{D}) \right)$  \COMMENT{\cref{eq:rep-loss}}
        \STATE Update critic
        \STATE \quad \quad $\psi \leftarrow \psi + \alpha \nabla \left( \mathcal{L}_{Q}(\psi; \mathcal{D}) \right)$ \COMMENT{\cref{eq:value-loss}}
        \IF{i \% 2 == 0}
          \STATE Update actor less frequently than critic
          \STATE $\quad \quad \eta \leftarrow \psi + \alpha \nabla \left( \mathcal{L}_{\pi}(\psi; \mathcal{D}) \right)$  \COMMENT{\cref{eq:policy-loss}}
          % \STATE $\bar{\eta} \leftarrow \tau \bar{\eta} + (1-\tau) \eta$
        \ENDIF
        \STATE Update target networks
        \STATE $\quad \quad [\bar{\theta}, \bar{\phi}, \bar{\psi}] \leftarrow \tau [\bar{\theta}, \bar{\phi}, \bar{\psi}] + (1-\tau) [{\theta}, {\phi}, {\psi}]$
    \ENDFOR
   \ENDFOR
\end{algorithmic}
\end{algorithm}

@@

\begin{table}[t]
\caption{FSQ levels $\mathcal{L}$ to approximate different codebook sizes $|\mathcal{C}|$}
\label{sample-table}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lccc}
\toprule
Target size $|\mathcal{C}|$ & $2^{8}$ & $2^{10}$ & $2^{12}$ \\
\midrule
Proposed $\mathcal{L}$ & $[8,6,5]$ & $[8,5,5,5]$ & $[7,5,5,5,5]$ \\
% \bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

\textbf{Method overview}
\our has four main components which we wish to learn:
\begin{align}
&\text{Encoder: } & z_{t} &= e_{\theta} (s_{t}) \label{eq:encoder} \\
&\text{Dynamics: } & \hat{z}_{t+1} &= d_{\phi} (z_{t}, a_{t}) \label{eq:transition} \\
&\text{Value: } & q_{t} &= Q_{\psi} (z_{t}, a_{t}) \label{eq:value} \\
&\text{Policy: } & a_{t} &\sim \pi_{\eta} (z_{t}) . \label{eq:policy}.
\end{align}
The encoder $e_{\theta}$  and latent-space dynamics model $d_{\phi}$ are responsible for representation learning.
The encoder $e_{\theta}$ maps observations $o_{t}$ to latent states $z_{t}$ and is responsible for learning a representation
which can aid RL.
The latent-space dynamics model $d_{\phi}$ predicts next latent states $z_{t+1}$ given a latent state $z_{t}$ and an action $a_{t}$.
It's sole purpose is to aid representation learning by making the latent states temporally consistent.
Note that we do not use it for model-based RL.
Once we have the representation learned by our encoder, we map all observations to the latent space and perform model-free RL
in this latent space.
Throughout this paper, we use Twin Delayed Deep Deterministic Policy Gradient
(TD3), \citep{fujimotoAddressingFunctionApproximation2018}).
It consists of two state-action value functions $\{Q_{\psi_{1}},Q_{\psi_{2}} \}$, known as critics, and a deterministic
actor $\pi_{\eta}$.
Following prior works \citep{yaratsMasteringVisualContinuous2021} we use a linear exploration noise schedule
which decays from $1$ to $0.1$ during training.

@@

\textbf{Representation learning}
Our representation learning is based solely on the latent-state consistency loss,
%
\begin{align} \label{eq:rep-loss}
  \mathcal{L}_{\text{rep}}(\theta, \phi; \tau)
 % &= \E_{s_{t} \sim \mathcal{D}} \left[ \sum_{h=0}^{H-1} \| \hat{z}_{t+1} - z_{t+1} \|_{2}^{2} \right] \\
%&= \E_{(s_{t}, a_{t}, s_{t+1}) \sim \mathcal{D}}
% &= \E_{\tau \sim \mathcal{D}}
% \left[ \sum_{h=0}^{H-1} \gamma^{h} \| d_{\phi}(e_{\theta}(s_{t}), a_{t}) - e_{\bar{\theta}}(s_{t+1}) \|_{2}^{2} \right]
&= \sum_{h=0}^{H-1} \gamma^{h} \| d_{\phi}(e_{\theta}(s_{t}), a_{t}) - e_{\bar{\theta}}(s_{t+1}) \|_{2}^{2},
\end{align}
%
which minimizes the distance between the next state predicted by the transition model $d_{\phi}(e_{\phi}(z_{t}), a_{t})$
and the next state predicted by the momentum encoder $e_{\bar{\phi}}(z_{t+1})$.
The prediction $d_{\phi}(e_{\phi}(z_{t}), a_{t})$ uses the online encoder which is being trained jointly with the transition model.
The target $e_{\bar{\phi}}(z_{t+1})$ is calculated with the momentum encoder which uses an exponential moving average (EMA)
of the encoder's weights $\bar{\theta} \leftarrow \tau \bar{\theta} + (1-\tau)\theta$.
The target network update rate is denoted $\tau$.

Unlike previous works, we normalize the latent vector

\textbf{Finite Scalar Quantization}
We follow Finite Scalar Quantization (FSQ, \cite{mentzerFiniteScalarQuantization2023})
and bound each dimension of our latent space.
Given the output of our MLP encoder $z \in \R^{d}$,

where each entry in  $f : z \rightarrow \left[ L/2 \right] \mathrm{tanh}(z))$

\textbf{Bounding the latent space}
Typically the latent state $z_{t} \in \R^{d}$
However, in our work we bound the domain of the latent space, \ie $z_{t} \in [-L, L]^{d}$.

@@

\section{Conclusion}
\label{conclusion}

>>>>

\begin{algorithm}[tb]
   \caption{\our}
   \label{alg:main_alg}
   \renewcommand{\algorithmiccomment}[1]{\hfill\textcolor{gray}{\(\triangleright\) #1}}
\begin{algorithmic}
   \STATE {\bfseries Input:} Encoder $e_{\theta}$, transition $d_{\phi}$, critics $\{Q_{\psi_{1}}, Q_{\psi_{2}} \}$, policy $\pi_{\eta}$, learning rate $\alpha$, target network update rate $\tau$
   \FOR{$i$ {\bfseries to} $N_{\text{episodes}}$}
    \STATE Collect data in environment:
    \STATE $\qquad \mathcal{D} \leftarrow \mathcal{D} \cup \{o_{t}, a_{t}, o_{t+1}, r_{t+1}\}^{T}_{t=0}$
    \FOR{$i=1$ {\bfseries to} $T$}
        \STATE Update representation:
        \STATE $\qquad [\theta, \phi] \leftarrow [\theta, \phi] + \alpha \nabla \left( \mathcal{L}_{\text{rep}}(\theta, \phi; \mathcal{D}) \right)$  \COMMENT{\cref{eq:rep-loss}}
        \STATE Update critic
        \STATE \quad \quad $\psi \leftarrow \psi + \alpha \nabla \left( \mathcal{L}_{Q}(\psi; \mathcal{D}) \right)$ \COMMENT{\cref{eq:value-loss}}
        \IF{i \% 2 == 0}
          \STATE Update actor less frequently than critic:
          \STATE $\quad \quad \eta \leftarrow \psi + \alpha \nabla \left( \mathcal{L}_{\pi}(\psi; \mathcal{D}) \right)$  \COMMENT{\cref{eq:policy-loss}}
          % \STATE $\bar{\eta} \leftarrow \tau \bar{\eta} + (1-\tau) \eta$
        \ENDIF
        \STATE Update target networks:
        \STATE $\quad \quad [\bar{\theta}, \bar{\phi}, \bar{\psi}] \leftarrow \tau [\bar{\theta}, \bar{\phi}, \bar{\psi}] + (1-\tau) [{\theta}, {\phi}, {\psi}]$
    \ENDFOR
   \ENDFOR
\end{algorithmic}
\end{algorithm}

@@

\textbf{Method overview}
\our has four main components which we wish to learn:
\begin{align}
&\text{Encoder: } & z_{t} &= e_{\theta} (s_{t}) \label{eq:encoder} \\
&\text{Dynamics: } & \hat{z}_{t+1} &= d_{\phi} (z_{t}, a_{t}) \label{eq:transition} \\
&\text{Value: } & q_{t} &= Q_{\psi} (z_{t}, a_{t}) \label{eq:value} \\
&\text{Policy: } & a_{t} &\sim \pi_{\eta} (z_{t}) . \label{eq:policy}
\end{align}
The encoder $e_{\theta}$  and latent-space dynamics model $d_{\phi}$ are responsible for representation learning.
The encoder $e_{\theta}$ maps observations $o_{t}$ to latent states $z_{t}$ and is responsible for learning a representation
which can aid RL.
The latent-space dynamics model $d_{\phi}$ predicts next latent states $z_{t+1}$ given a latent state $z_{t}$ and an action $a_{t}$.
It's sole purpose is to aid representation learning by making the latent states temporally consistent.
Note that we do not use it for model-based RL.
Once we have the representation learned by our encoder, we map all observations to the latent space and perform model-free RL
in this latent space.
Throughout this paper, we use Twin Delayed Deep Deterministic Policy Gradient
(TD3), \citep{fujimotoAddressingFunctionApproximation2018}).
It consists of two state-action value functions $\{Q_{\psi_{1}},Q_{\psi_{2}} \}$, known as critics, and a deterministic
actor $\pi_{\eta}$.
Following prior works \citep{yaratsMasteringVisualContinuous2021} we use a linear exploration noise schedule
which decays from $1$ to $0.1$ during training.

@@

\textbf{Representation learning}
Our representation learning is based solely on the latent-state consistency loss,
%
\begin{align} \label{eq:rep-loss}
  \mathcal{L}_{\text{rep}}(\theta, \phi; \tau)
 % &= \E_{s_{t} \sim \mathcal{D}} \left[ \sum_{h=0}^{H-1} \| \hat{z}_{t+1} - z_{t+1} \|_{2}^{2} \right] \\
%&= \E_{(s_{t}, a_{t}, s_{t+1}) \sim \mathcal{D}}
% &= \E_{\tau \sim \mathcal{D}}
% \left[ \sum_{h=0}^{H-1} \gamma^{h} \| d_{\phi}(e_{\theta}(s_{t}), a_{t}) - e_{\bar{\theta}}(s_{t+1}) \|_{2}^{2} \right]
&= \sum_{h=0}^{H-1} \gamma^{h} \| d_{\phi}(e_{\theta}(s_{t}), a_{t}) - e_{\bar{\theta}}(s_{t+1}) \|_{2}^{2},
\end{align}
%
which minimizes the distance between the next state predicted by the transition model $\hat{z}_{t+1} = d_{\phi}(e_{\phi}(z_{t}), a_{t})$
and the next state predicted by the momentum encoder $\bar{z}_{t+1} = e_{\bar{\phi}}(z_{t+1})$.
The prediction $d_{\phi}(e_{\phi}(z_{t}), a_{t})$ uses the online encoder which is being trained jointly with the transition model.
The target $e_{\bar{\phi}}(z_{t+1})$ is calculated with the momentum encoder which uses an exponential moving average (EMA)
of the encoder's weights $\bar{\theta} \leftarrow \tau \bar{\theta} + (1-\tau)\theta$.
% As such, gradient to not flow through the target network
The target network update rate is denoted $\tau$.

\begin{table}[t]
\caption{FSQ levels $\mathcal{L}$ to approximate different codebook sizes $|\mathcal{C}|$}
\label{tab:fsq-levels}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lccc}
\toprule
Target size $|\mathcal{C}|$ & $2^{6}$ & $2^{8}$ & $2^{10}$ \\
\midrule
Proposed $\mathcal{L}$ & $[8,8]$ & $[8,6,5]$ & $[8,5,5,5]$ \\
% Target size $|\mathcal{C}|$ & $2^{8}$ & $2^{10}$ & $2^{12}$ \\
% \midrule
% Proposed $\mathcal{L}$ & $[8,6,5]$ & $[8,5,5,5]$ & $[7,5,5,5,5]$ \\
% \bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

\textbf{Normalization with implicit quantization}
% However, this can lead to exploding gradients and
Motivated by preventing representation collapse we seek a representation with high matrix rank.
To this end, we bound our latent space following the approach from
Finite Scalar Quantization \citep[FSQ,][]{mentzerFiniteScalarQuantization2023}.
Note that the latent state $z_{t}$ is typically a vector of unbounded real numbers, \ie $z_{t} \in \R^{d}$.
% each (scalar) entry in the representation $z_{t}$ is independently quantized to the nearest integer by rounding
Their important observation is that carefully bounding each channel gives rise to an implicit codebook of a chosen size.
Let's consider a single element of the latent state $z$ and give it $c$ channels so it's of shape $[c]$.
% Here $d$ is the output dimension of the NN.
Mapping each entry $z_{i}$ to $L$ values (\eg, via $z_{i} \rightarrow [L/2] \text{tanh}(z_{i})$ and rounding to integers),
obtains a quantized $\tilde{z} \in \mathcal{C}$, where $\tilde{z}$ takes one of $L^{c}$ unique possible vectors.
This gives rise to an implicit codebook $\mathcal{C}$ given by the product of the per-channel codebook sets.
% The size of the codebook $|\mathcal{C}| = L^{c}$ is given by the product of the
% For example, specifying $c=3$ channels
This idea generalizes to the case where the $i^{\text{th}}$ channel is mapped to $L_{i}$ values.
Giving a codebook of size $|\mathcal{C}| = L_{i}^{d_{i}}$.
% Since quantization is performed by rounding to integers

FSQ has the following hyperparameters. We must specify the number of channels $c$ and the number of levels per channel
$\mathcal{L} = [L_{1},\ldots,L_{d}]$.
\cref{tab:fsq-levels} shows the recommended number of channels and number of levels per channel to obatin codebooks of
different sizes.
FSQ, relies on a simple, fixed grid partition in a much
lower-dimensional space.
In practice, we found codebooks of size $|\mathcal{C}|=2^{6}$ sufficient for most environments in the DeepMind Control suite.
However, for the dog-walk environment we used a codebook of size $|\mathcal{C}|=2^{12}$.

% Fig. 1 shows FSQ for d=3, L=3, implying a codebook
% C = {(−1, −1, −1),(−1, −1, 0),(−1, −1, 1), . . . ,(1, 1, 1)}, where |C| = L
% d = 27.

% However, thi w
% bounding the range of the quantizer
% However, in our work we bound the domain of the latent space, \ie $z_{t} \in [-L, L]^{d}$.
% We follow Finite Scalar Quantization (FSQ, \cite{mentzerFiniteScalarQuantization2023})
% and bound each dimension of our latent space.
% Given the output of our MLP encoder $z \in \R^{d}$,

% where each entry in  $f : z \rightarrow \left[ L/2 \right] \mathrm{tanh}(z))$

% Unlike previous works, we normalize the latent vector

% \textbf{Bounding the latent space}

@@

\section{Conclusion}
\label{conclusion}
We have presented \our, a simple technique for learning task-agnostic representations which can be combined with
any model-free RL algorithm.

\textbf{Future work}
To the best of our knowledge, we are the first to present a method that learns a task-agnostic representation
using only latent-state consistency. As such, we hope our method for learning task-agnostic representations paves the
way for more research in multi-task and lifelong learning.
Given that our method obtains an implicit codebook an obvious future direction is exploring if we can leverage the
discretized latent-space.

\textbf{Limitations}
